---
title: "Initial QC on RD7 Essential Data - Multi-UTR Genes"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

Here we perform a second pass look at the data, similar to the first pass, but
focusing specifically on multi-UTR genes.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(plotly)
library(ggbeeswarm)
library(Matrix)
library(matrixStats)
library(SingleCellExperiment)
```

## Parameters
```{r set_params}
set.seed(20210818)
FILE_SCE_TX="data/sce/kd6_essential.annot.txs.Rds"

## KNOWN FACTORS
GENES_CPSF <- c("CPSF1", "CPSF2", "CPSF3", "CPSF4", "WDR33", "FIP1L1")
GENES_CF1 <- c("NUDT21", "CPSF6", "CPSF7")
GENES_CF2 <- c("PCF11", "CLP1")
GENES_CSTF <- c("CSTF1", "CSTF2", "CSTF2T", "CSTF3")
GENES_PAF <- c("PAF1", "CTR9", "RTF1")
GENES_NEF <- c("NXF1", "THOC1", "THOC2", "THOC3", "THOC5", "THOC6", "THOC7")
GENES_SPL <- str_c("SRSF", 1:11)

COLOR_MAP <- c("non-targeting"="grey70",
               "CPSF complex"="palegreen2",
               "CFI complex"="plum3",
               "CFII complex"="orange2",
               "CSTF complex"="peachpuff",
               "PAF complex"="steelblue",
               "mRNA export factors"="lightcoral",
               "splicing factors"="cyan3",
               "other"="black")
```

# Data
## Loading
```{r load_data, message=FALSE}
sce <- readRDS(FILE_SCE_TX) %>% `[`(rowData(.)$utr_type_raw == 'multi',)
```

## Preprocessing
Among our statistics, we will compute the percentage of UMIs corresponding to specific
isoform classes, including IPA, proximal, and distal, in multi-UTR genes.

```{r prepare_data}
df_cells <- colData(sce) %>% as_tibble %>%
    mutate(umis_total=colSums(counts(sce)),
           pct_ipa=colSums(counts(sce[rowData(sce)$is_ipa,]))/umis_total,
           pct_proximal=colSums(counts(sce[rowData(sce)$is_proximal,]))/umis_total,
           pct_distal=colSums(counts(sce[rowData(sce)$is_distal,]))/umis_total)
```

# Target Analysis
We will explore whether any perturbations have effects on basic transcriptome characteristics.
These will also be compared against each other.

```{r compute_target_stats}
df_targets <- df_cells %>%
    group_by(target_gene, target_gene_id, sgID_AB) %>%
    summarize(n_cells=dplyr::n(),
              mean_umis_total=mean(umis_total),
              pct_ipa=weighted.mean(pct_ipa, umis_total),
              pct_proximal=weighted.mean(pct_proximal, umis_total),
              pct_distal=weighted.mean(pct_distal, umis_total), .groups='drop') %>%
    mutate(gene_set=case_when(
        target_gene == "non-targeting" ~ "non-targeting",
        target_gene %in% GENES_CPSF ~ "CPSF complex",
        target_gene %in% GENES_CF1 ~ "CFI complex",
        target_gene %in% GENES_CF2 ~ "CFII complex",
        target_gene %in% GENES_CSTF ~ "CSTF complex",
        target_gene %in% GENES_PAF ~ "PAF complex",
        target_gene %in% GENES_NEF ~ "mRNA export factors",
        target_gene %in% GENES_SPL ~ "splicing factors",
        TRUE ~ "other"
    ) %>% factor(levels=c("other", "non-targeting", "CPSF complex", 
                          "CFI complex", "CFII complex", "CSTF complex",
                          "PAF complex", "mRNA export factors",
                          "splicing factors"))) %>%
    mutate(is_known=!gene_set %in% c("non-targeting", "other"))
```


### Proximal Extemes
```{r proximal_min_max}
df_targets %>%
    select(target_gene, n_cells, pct_proximal, mean_umis_total, sgID_AB) %>%
    slice_max(pct_proximal, n=20)

df_targets %>%
    select(target_gene, n_cells, pct_proximal, mean_umis_total, sgID_AB) %>%
    slice_min(pct_proximal, n=20)
```

### Distal Extremes
```{r distal_min_max}
df_targets %>%
    select(target_gene, n_cells, pct_distal, mean_umis_total, sgID_AB) %>%
    slice_max(pct_distal, n=20)

df_targets %>%
    select(target_gene, n_cells, pct_distal, mean_umis_total, sgID_AB) %>%
    slice_min(pct_distal, n=20)
```

## Pair Plots

## Proximal vs Distal
```{r plt_pair_pr_ds_static, fig.width=8, fig.height=6}
df_targets %>%
    ggplot(aes(x=pct_proximal, y=pct_distal, text=target_gene)) +
    geom_point(aes(size=n_cells), data=filter(df_targets, gene_set == 'other')) +
    geom_point(data=filter(df_targets, gene_set == 'non-targeting'), size=0.7, color='grey') +
    geom_point(aes(fill=gene_set), data=filter(df_targets, is_known), size=3, pch=21, color='black') +
    geom_rug(data=filter(df_targets, gene_set != 'non-targeting'), 
             size=0.05, length=unit(0.02, "npc")) +
    geom_rug(data=filter(df_targets, gene_set == 'non-targeting'), 
             size=0.1, length=unit(0.04, "npc"), color='grey') +
    scale_fill_manual(values=COLOR_MAP) +
    scale_size_continuous(range=c(0.005, 1), guide=NULL) +
    scale_x_continuous(labels=scales::percent_format(accuracy=0.1)) +
    scale_y_continuous(labels=scales::percent_format(accuracy=0.1)) +
    theme_bw() +
    labs(x="Percent UMIs from proximal isoforms",
         y="Percent UMIs from distal isoforms",
         fill="Gene Set")
```


```{r export_plt_pr_ds}
ggsave("img/su-lu-known-factors-kd6.pdf", width=8.2, height=6, dpi=300)
```

## IPA vs Proximal
```{r plt_pair_ipa_pr_static, fig.width=8, fig.height=6}
df_targets %>%
    ggplot(aes(x=pct_ipa, y=pct_proximal, text=target_gene)) +
    geom_point(aes(size=n_cells), data=filter(df_targets, gene_set == 'other')) +
    geom_point(data=filter(df_targets, gene_set == 'non-targeting'), size=0.7, color='grey') +
    geom_point(aes(fill=gene_set), data=filter(df_targets, is_known), size=3, pch=21, color='black') +
    geom_rug(data=filter(df_targets, gene_set != 'non-targeting'), 
             size=0.05, length=unit(0.02, "npc")) +
    geom_rug(data=filter(df_targets, gene_set == 'non-targeting'), 
             size=0.1, length=unit(0.04, "npc"), color='grey') +
    scale_fill_manual(values=COLOR_MAP) +
    scale_size_continuous(range=c(0.005, 1), guide=NULL) +
    scale_x_continuous(labels=scales::percent_format(accuracy=0.1)) +
    scale_y_continuous(labels=scales::percent_format(accuracy=0.1)) +
    theme_bw() +
    labs(x="Percent UMIs from IPA isoforms",
         y="Percent UMIs from proximal isoforms",
         fill="Gene Set")
```

```{r export_plt_ipa_pr}
ggsave("img/ipa-su-known-factors-kd6.pdf", width=8.2, height=6, dpi=300)
```


## IPA vs Distal
```{r plt_pair_ipa_ds_static, fig.width=8, fig.height=6}
df_targets %>%
    ggplot(aes(x=pct_ipa, y=pct_distal, text=target_gene)) +
    geom_point(aes(size=n_cells), data=filter(df_targets, gene_set == 'other')) +
    geom_point(data=filter(df_targets, gene_set == 'non-targeting'), size=0.7, color='grey') +
    geom_point(aes(fill=gene_set), data=filter(df_targets, is_known), size=3, pch=21, color='black') +
    geom_rug(data=filter(df_targets, gene_set != 'non-targeting'), 
             size=0.05, length=unit(0.02, "npc")) +
    geom_rug(data=filter(df_targets, gene_set == 'non-targeting'), 
             size=0.1, length=unit(0.04, "npc"), color='grey') +
    scale_fill_manual(values=COLOR_MAP) +
    scale_size_continuous(range=c(0.005, 1), guide=NULL) +
    scale_x_continuous(labels=scales::percent_format(accuracy=0.1)) +
    scale_y_continuous(labels=scales::percent_format(accuracy=0.1)) +
    theme_bw() +
    labs(x="Percent UMIs from IPA isoforms",
         y="Percent UMIs from distal isoforms",
         fill="Gene Set")
```

```{r export_plt_ipa_ds}
ggsave("img/ipa-ds-known-factors-kd6.pdf", width=8.2, height=6, dpi=300)
```


# Conclusion

Considering on multi-UTR genes provides a cleaner signal for effects on different
isoforms. We now see many of the known factors stand out, and can identify other
possible genes with isoform-specific effects.

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
