---
title: "Ordered Clusters of KD6 Essential Data"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

Final heatmaps for the K562 6 Day data.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(pheatmap)
library(bluster)
library(Matrix)
```

## Parameters
```{r set_params}
set.seed(20210818)

CLUSTERS_GENE_UNREG=c(9,12,13)

## input files
FILE_CLUSTERS_TARGETS="tbl/df_target_nnclusters_kd6_essential_ui10.csv"
FILE_CLUSTERS_GENES="tbl/df_gene_nnclusters_kd6_essential_ui10.csv"
FILE_REORDER="metadata/kd6-cluster-info-sm.csv"
RDS_DWUI="data/dwui/20221107-dwui-target-gene-kd6-essential.Rds"
RDS_ZDWUI="data/dwui/20221107-zdwui-target-gene-kd6-essential.Rds"

## output files
FILE_OUT_AVG_TARGET="img/heatmap-kd6-essential-avg-target-clusters-final.pdf"
FILE_OUT_AVG_GENE="img/heatmap-kd6-essential-avg-gene-clusters-final.pdf"
FILE_OUT_AVG_BOTH="img/heatmap-kd6-essential-avg-both-clusters-final.pdf"
FILE_OUT_FINAL="img/heatmap-kd6-essential-nnclusters-final.pdf"

## aesthetics
NCOLORS=100
COLORS_BWR <- colorRampPalette(c("blue", "white", "red"))(NCOLORS)
COLORS_MKY <- colorRampPalette(c("magenta", "black", "yellow"))(NCOLORS)
COLORS_YKM <- colorRampPalette(c("yellow", "black", "magenta"))(NCOLORS)
COLORS_RKG <- colorRampPalette(c("red", "black", "green"))(NCOLORS)
breaks_dwui <- seq(-0.5, 0.5, length.out=NCOLORS + 1)
breaks_zdwui <- seq(-4, 4, length.out=NCOLORS + 1)
breaks_zdwui_broad <- seq(-6, 6, length.out=NCOLORS + 1)
breaks_pca <- seq(-20, 20, length.out=NCOLORS + 1)
```

# Data
## Loading
```{r load_data, message=FALSE}
dwui_target_gene <- readRDS("data/dwui/20221107-dwui-target-gene-kd6-essential.Rds")
dwui_target_gene[is.na(dwui_target_gene)] <- 0
zdwui_target_gene <- readRDS("data/dwui/20221107-zdwui-target-gene-kd6-essential.Rds")
zdwui_target_gene[is.na(zdwui_target_gene)] <- 0

map_cluster_order <- read_csv(FILE_REORDER, col_select=c("cluster_id_target","order_heatm_new")) %>% deframe()
df_cluster_target <- read_csv(FILE_CLUSTERS_TARGETS) %>% 
    ## apply renaming
    mutate(cluster_id=factor(map_cluster_order[as.character(cluster_id)], 
                             levels=sort(unique(map_cluster_order))))
df_cluster_gene <- read_csv(FILE_CLUSTERS_GENES) %>% 
    mutate(cluster_id=factor(cluster_id, levels=sort(unique(cluster_id))))
```

## Preprocessing

```{r prepare_data}
sgid2gene <- df_cluster_target %>%
    dplyr::select(sgID_AB, target_gene) %>%
    distinct(sgID_AB, target_gene) %>%
    deframe()

ens2gene <- df_cluster_gene %>%
    dplyr::select(gene_id, gene_name) %>%
    distinct(gene_id, gene_name) %>%
    deframe()

convert_rownames <- function (mat, in2out) {
    mat %>% set_rownames(in2out[rownames(.)])
}

convert_colnames <- function (mat, in2out) {
    mat %>% set_colnames(in2out[colnames(.)])
}

idx_targets1 <- df_cluster_target %>%
    arrange(cluster_id, target_gene) %$%
    sgID_AB

idx_genes1 <- df_cluster_gene %>%
    arrange(cluster_id, gene_name) %$%
    gene_id

idx_genes_clean <- df_cluster_gene %>%
    filter(!(cluster_id %in% CLUSTERS_GENE_UNREG)) %>%
    arrange(cluster_id, gene_name) %$%
    gene_id
```

# Compute Cluster Means
## Target Clusters
```{r compute_target_cluster_means}
I_cluster_target <- df_cluster_target %>%
    dplyr::select(sgID_AB, cluster_id) %>%
    deframe %>% fac2sparse

## cluster target means
M_cluster_target <- I_cluster_target %>% { . / rowSums(.) }

mzdwui_cluster_gene <- M_cluster_target[,idx_targets1] %*% zdwui_target_gene[idx_targets1,]
mzdwui_cluster_gene %<>% as.matrix
```

### Plot Target Cluster Means
```{r plt_hc_cluster_mzdwui, fig.width=8, fig.height=4}
df_col_annots <- df_cluster_gene %>%
    dplyr::select(gene_id, cluster_id) %>%
    rename(gene_cluster=cluster_id) %>%
    column_to_rownames("gene_id")

pheatmap(mzdwui_cluster_gene[, idx_genes_clean], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=1, fontsize_row=10,
         annotation_col=df_col_annots,
         show_colnames=FALSE, show_rownames=TRUE, scale='none', 
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete",
         cutree_rows=NA, cutree_cols=NA)
```

### Reordered Target Clusters DWUI
```{r plt_ordered_cluster_dwui, fig.width=8, fig.height=8}
idx_targets2 <- df_cluster_target %>%
    arrange(cluster_id, target_gene) %$%
    sgID_AB

df_row_annots <- df_cluster_target %>%
    dplyr::select(sgID_AB, cluster_id) %>%
    rename(target_cluster=cluster_id) %>%
    column_to_rownames("sgID_AB")

pheatmap(zdwui_target_gene[idx_targets2, idx_genes_clean], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=1, fontsize_row=10,
         annotation_row=df_row_annots, 
         annotation_col=df_col_annots,
         show_colnames=FALSE, show_rownames=FALSE, scale='none',
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete")
```

## Gene Clusters
```{r compute_gene_cluster_means}
I_cluster_gene <- df_cluster_gene %>%
    filter(!(cluster_id %in% CLUSTERS_GENE_UNREG)) %>%
    dplyr::select(gene_id, cluster_id) %>%
    deframe %>% fac2sparse

## cluster target means
M_cluster_gene <- I_cluster_gene %>% { . / rowSums(.) }

mzdwui_target_cluster <- zdwui_target_gene[,idx_genes_clean] %*% t(M_cluster_gene[,idx_genes_clean])
mzdwui_target_cluster %<>% as.matrix
```

### Plot Gene Cluster Means
```{r plt_hc_cluster_gene_mzdwui, fig.width=5, fig.height=6}
pheatmap(mzdwui_target_cluster[idx_targets2,], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=10, fontsize_row=10,
         show_colnames=TRUE, show_rownames=FALSE, scale='none', 
         angle_col=0, annotation_names_row=FALSE,
         annotation_row=df_row_annots, 
         cluster_rows=FALSE, cluster_cols=TRUE, clustering_method="complete", 
         drop_levels=TRUE)
```

### Reordered Gene Cluster Means
```{r plt_ordered_cluster_gene_mzdwui, fig.width=4, fig.height=8}
idx_clusters_genes <- as.character(c(20,16,7,8,1,3,18,2,5,10,15,19,14,11,4,6,17))

pheatmap(mzdwui_target_cluster[idx_targets2, idx_clusters_genes], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=10, fontsize_row=10,
         show_colnames=TRUE, show_rownames=FALSE, scale='none', 
         annotation_row=df_row_annots, 
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete", 
         drop_levels=TRUE)
```

### Reordered All Clusters DWUI
```{r plt_ordered_clusters_all_dwui, fig.width=8, fig.height=8}
idx_genes2 <- df_cluster_gene %>%
    filter(!(cluster_id %in% CLUSTERS_GENE_UNREG)) %>%
    mutate(cluster_id=factor(cluster_id, levels=idx_clusters_genes)) %>%
    arrange(cluster_id, gene_name) %$%
    gene_id

df_col_annots <- df_cluster_gene %>%
    dplyr::select(gene_id, cluster_id) %>%
    mutate(cluster_id=factor(cluster_id, levels=idx_clusters_genes)) %>%
    rename(gene_cluster=cluster_id) %>%
    column_to_rownames("gene_id")

gaps_col <- df_cluster_gene %>%
    filter(!(cluster_id %in% CLUSTERS_GENE_UNREG)) %>%
    mutate(cluster_id=factor(cluster_id, levels=idx_clusters_genes)) %>%
    arrange(cluster_id, gene_name) %$%
    table(cluster_id) %>%
    cumsum

gaps_row <- df_cluster_target %>%
    arrange(cluster_id, target_gene) %$%
    table(cluster_id) %>%
    cumsum
         
pheatmap(zdwui_target_gene[idx_targets2, idx_genes2], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=1, fontsize_row=10,
         annotation_row=df_row_annots, 
         annotation_col=df_col_annots,
         show_colnames=FALSE, show_rownames=FALSE, scale='none',
         annotation_names_row=FALSE, annotation_names_col=FALSE,
         gaps_row=gaps_row,
         gaps_col=gaps_col,
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete")
```

# Final Joint Plot
```{r joint_plot, figure.width=16, fig.height=16}
pheatmap(zdwui_target_gene[idx_targets2, idx_genes2], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=1, fontsize_row=1,
         annotation_row=df_row_annots, 
         annotation_col=df_col_annots,
         show_colnames=TRUE, show_rownames=TRUE, scale='none',
         labels_row=sgid2gene[idx_targets2],
         labels_col=ens2gene[idx_genes2],
         annotation_names_row=FALSE, annotation_names_col=FALSE,
         gaps_row=gaps_row,
         gaps_col=gaps_col,
         cluster_rows=FALSE, cluster_cols=FALSE,
         filename=FILE_OUT_FINAL, width=16, height=16)
```


#### Export Clustered Targets
```{r export_avg_target_dendro}
pheatmap(mzdwui_cluster_gene[, idx_genes2], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=1, fontsize_row=10,
         annotation_col=df_col_annots,
         annotation_names_row=FALSE, annotation_names_col=FALSE,
         show_colnames=FALSE, show_rownames=TRUE, scale='none', 
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete",
         cutree_rows=NA, cutree_cols=NA, 
         filename=FILE_OUT_AVG_TARGET, width=8, height=4)
```

#### Export Clustered Genes
```{r export_avg_gene_dendro}
pheatmap(mzdwui_target_cluster[idx_targets2,], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=10, fontsize_row=10,
         show_colnames=TRUE, show_rownames=FALSE, scale='none', 
         angle_col=0,
         annotation_row=df_row_annots, 
         annotation_names_row=FALSE, annotation_names_col=FALSE,
         cluster_rows=FALSE, cluster_cols=TRUE, clustering_method="complete", 
         drop_levels=TRUE, 
         filename=FILE_OUT_AVG_GENE, width=5, height=6)
```


### Double Average
```{r plt_double_average, fig.width=5, fig.height=4}
mzdwui_cluster_cluster <- M_cluster_target[,idx_targets2] %*% mzdwui_target_cluster[idx_targets2,]

pheatmap(mzdwui_cluster_cluster[,idx_clusters_genes], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=10, fontsize_row=10,
         show_colnames=TRUE, show_rownames=TRUE, scale='none', 
         angle_col=0,
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete")
```

### Export Double Average
```{r export_double_average, fig.width=5, fig.height=4}
pheatmap(mzdwui_cluster_cluster[,idx_clusters_genes], 
         color=COLORS_BWR,
         breaks=breaks_zdwui,
         fontsize_col=10, fontsize_row=10,
         show_colnames=TRUE, show_rownames=TRUE, scale='none', 
         angle_col=0,
         cluster_rows=FALSE, cluster_cols=FALSE, clustering_method="complete",
         filename=FILE_OUT_AVG_BOTH, width=5, height=4)
```


---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
